Highly Available Transactions: Virtues and Limitations

### 2013-03-19 writeup/hat-analysis.tex
More writing

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. point out, Gray's original Repeatable Read lock-based
implementation provides substantially richer guarantees than those
that are promised by the ANSI SQL specification. Gray, Berenson, and
Adya effectively preclude all serializability anomalies but the
Phantom Problem (which we have explicitly not yet considered and will
do so in Section~\ref{sec:unachievable-hat}). Nevertheless, the ANSI
SQL specification provides a useful property which, although is not
true to Gray's spirit of Repeatable Read, is used in distributed
consistency models. The ANSI Repeatable Read guarantee requires that,
along with observing Read Committed isolation, each transaction can
only read only version of each data item that it did not itself
produce. In the example below, $T_3$'s must see $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is much weaker than serializability, but it is
a good faith literal interpretation of the phrase ``Repeatable Read'':
transactions see their own writes. By itself, this property can be
implemented trivially---always read \texttt{null} for each
key. However, when coupled with additional properties, like
transactional atomicity, it becomes stronger: the repeated reads must
also obey some additional ordering properties. Distributed systems
describe a ``consistent snapshot'' across a set of related events as a
\textit{cut} across a set of participants or, in our case, data
items. Accordingly, to capture the notion that the transaction should
read from a consistent cut, and to disambiguate from the
aforementioned stronger Repeatable Read properties, we call ANSI SQL
Repeatable Read \textit{cut isolation}. In the absence of additional
guarantees, it is possible to satisfy cut isolation with high
availability by caching each transaction's reads or, alternatively,
using multi-versioning on the server and ensuring that each
transaction's successive reads return the same version of each data
item.

@@

@@

Finally, to require that servers propagate writes between replicas, we
can require \textbf{convergence}, or eventual consistency for each
data item: in the absence of new mutations to a data item, in the
absence of partitions, all servers eventually agree on the value for
each item. This is typically accomplished by any number of
anti-entropy protocols, which periodically update neighboring servers
with the latest value for each data item.

@@

* Lost Update
	* Simple proof: T1 R(x) W(x=x+1) T2 R(x) W(x=x+1); execute on opposite sides 
* Read Skew
	* Similar easy proof
* Write Skew
	* Similar easy proof
* Anti-dependency cycles
	* Serializability
* Recency guarantees on data items
	* CAP

@@

 \newcommand{\lostupdate}{$^\dagger$}
 \newcommand{\rwskew}{$^\ddagger$}
 \newcommand{\mphantom}{$^\star$}
 \newcommand{\linearizable}{$^\oplus$}

\begin{table}
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Cut Isolation (CI), Monotonic Reads (MR), Monotonic
Writes (MW), Writes Follow Reads (WFR)\\\hline Sticky-HA & Read Your
Writes (RYW), Causal Consistency\\\hline Unavailable & Cursor
Stability (CS)\lostupdate, Read Consistency (CR)\lostupdate, Snapshot
Isolation (SI)\lostupdate, Repeatable Read (RR)\lostupdate\rwskew,
Serializability (S)\lostupdate\rwskew\mphantom, Causal
Serializability\lostupdate\rwskew\mphantom, Strict
Serializability\lostupdate\rwskew\mphantom\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing read/write skew\rwskew, preventing
  phantoms\mphantom, and requiring linearizability\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \node[draw=none,fill=none] (ci) at (1.2, 1.6) {\underline{CI}};
 \node[draw=none,fill=none] (rc) at (1.2, .8) {\underline{RC}};
 \node[draw=none,fill=none] (ru) at (1.2, 0) {\underline{RU}};

 \node[draw=none,fill=none] (ta) at (0, 0) {\underline{TA}};
 \node[draw=none,fill=none] (mr) at (2.4, 0) {\underline{MR}};
 \node[draw=none,fill=none] (mw) at (3.6, 0) {\underline{MW}};
 \node[draw=none,fill=none] (wfr) at (4.8,0) {\underline{WFR}};
 \node at (6,0) [sticky] (ryw) {RYW};

 \node at (3.6, 2) [sticky] (causal) {Causal};
 \node[noha] (cs) at (1.7, 2.4) {CS};
 \node[noha] (rr) at (.7, 2.6) {RR};
 \node[noha] (si) at (-.4, 2.2) {SI};
 \node[noha] (cr) at (-.4, 1.2) {CR};
 \node[noha] (s) at (.5, 3.5) {S};
 \node[noha] (causal-s) at (3.6, 3.5) {Causal S};
 \node[noha] (strict-s) at (6, 3.5) {Strict S};

 \draw [->] (ru) -- (rc);
 \draw [->] (rc) -- (ci);

@@

 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (mr);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (ta);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ci) -- (ta);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (mr);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ci) -- (mr);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ta) -- (ru);
 \draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (causal);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (mr) -- (mw);
 \draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (mw);
 \draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (ryw);

 \draw [->, red] (ci) -- (cs);
 \draw [->, red] (cs) -- (rr);
 \draw [->, red] (ci) -- (si);
 \draw [->, red] (rr) -- (s);
 \draw [->, red] (si) -- (s);
 \draw [->, red] (ta) -- (cr);
 \draw [->, red] (rc) -- (cr);
 \draw [->, red] (cr) -- (si);
 \draw [->, red] (ta) -- (rr);
 \draw [->, red] (s) -- (causal-s);
 \draw [->, red] (causal) -- (causal-s);
 \draw [->, red] (causal-s) -- (strict-s);

\end{tikzpicture}
\label{fig:hat-order}
\caption{HAT, sticky HAT (in boxes), and unavailable models (circled)
  from Table~\protect\ref{table:hatcompared} compared graphically. Directed
  edges represent ordering by model strength, squiggle edges represent
  (transitive) compatible models, while absence of edges represents
  incomparable models. The combination of all underlined models
  represents maximum ACID HAT semantics, while the combination of
  underlined and boxed models represents maximum ACID sticky HAT
  models.}
\label{fig:hatcompared}
\end{figure}

>>>>

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. point out, Gray's original Repeatable Read lock-based
implementation provides substantially richer guarantees than those
that are promised by the ANSI SQL specification. Gray, Berenson, and
Adya effectively preclude all serializability anomalies but the
Phantom Problem (which we have explicitly not yet considered and will
do so in Section~\ref{sec:unachievable-hat}). Nevertheless, the ANSI
SQL specification provides a useful property which, although is not
true to Gray's spirit of Repeatable Read, is used in distributed
consistency models. The ANSI Repeatable Read guarantee requires that,
along with observing Read Committed isolation, each transaction can
only read only version of each data item that it did not itself
produce (Berenson et al. call this preventing ``Fuzzy Read,'' or
P2). In the example below, $T_3$'s must see $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is much weaker than serializability, but it is
a good faith literal interpretation of the phrase ``Repeatable Read'':
transactions see their own writes. By itself, this property can be
implemented trivially---always read \texttt{null} for each
key. However, when coupled with additional properties, like
transactional atomicity, it becomes stronger: the repeated reads must
also obey some additional ordering properties. Distributed systems
describe a ``consistent snapshot'' across a set of related events as a
\textit{cut} across a set of participants or, in our case, data
items. Accordingly, to capture the notion that the transaction should
read from a consistent cut, and to disambiguate from the
aforementioned stronger Repeatable Read properties, we call ANSI SQL
Repeatable Read \textbf{Cut Isolation}. In the absence of additional
guarantees, it is possible to satisfy cut isolation with high
availability by caching each transaction's reads or, alternatively,
using multi-versioning on the server and ensuring that each
transaction's successive reads return the same version of each data
item.

We can consider two variants of Cut Isolation. The first is relatively
straightforward, and reasons about cuts over discrete data items. We
will call this \textit{Item Cut Isolation}. The second is known as
preventing the \textit{Phantom Problem}, whereby two successive
predicate-based reads return different data. We call predicate-based
cut isolation, which prevents Phansoms, \textbf{Phantom Cut
  Isolation}.

@@

Together with Cut Isolation, Transactional Atomicity prevents Read
Skew (Berenson et al.'s A5A).

@@

Finally, to require that servers propagate writes between replicas, we
can require \textbf{convergence}, or eventual consistency for each
data item: in the absence of new mutations to a data item, in the
absence of partitions, all servers eventually agree on the value for
each item. This is typically accomplished by any number of
anti-entropy protocols, which periodically update neighboring servers
with the latest value for each data item. This convergence property
should be considered mandatory given the spirit of existing literature
on isolation protocols: in all models we have encountered, the
database should behave as if there is some total order over
transactions. Indeed, non-convergent systems are remarkably difficult
to reason about.

@@

The simplest way to describe which isolation properties are achievable
with high availability is in terms of preventing different isolation
anomalies. We have already shown that it is possible prevent several
anomalies, like Dirty Read, Dirty Write, Fuzzy Read, and atomicity
violations. In this section, we demonstrate that preventing Lost
Update, Write Skew, and Phantoms requires unavailability.

In the words of Berenson et al., \textbf{Lost Update} occurs when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider the following transactions:
\begin{align*}
\small
T_1 &: r_x(a) w_x(a+1)
\\T_2 &: w_x(2)
\end{align*}
If $T_1$'s read from $x$ precedes $T_2$'s write to $x$ but $T_2$'s
write to $x$ precedes $T_1$'s write to $x$, then a database containing
only these two transactions will end in a state that could not have
resulted in a serial execution.

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)
\end{align*}
The final state of the database will converge to an execution that
could not have resulted from a serial excecution. To prevent this from
happening, either $T_1$ or $T_2$ should not have committed. Each
client's respective server might try to detect that another write
occurred, but this requires knowing the version of the latest write to
$x$. In this example, this reduces to a requirement for
linearizability, which is, via Gilbert and Lynch's proof of the CAP
Theorem, provably unachievable with high availabilty.

\textbf{Write Skew} is a generalization of Lost Update to multiple
keys. It occurs when one transaction $T1$ reads a given data item $x$,
a second transaction $T2$ reads a different data item $y$, then $T1$
writes to $y$ and commits and $T2$ writes to $x$ and commits. As an
example of Write Skew, consider the following two transactions:
\begin{align*}
\small
T_1 &: r_y(0)~w_x(1)
\\T_2 &: r_x(0)~w_y(1)
\end{align*}
As Berenson et al. describe, if there was an integrity constraint
between $x$ and $y$ such that only one of $x$ or $y$ should be one at
any given time, then this write skew would lead to a non-serializable
execution. Write skew in particular is a somewhat esoteric
anomaly---for example, it does not appear in TPC-C---but, as a
generalization of Lost Update, is also unavailable.

\textbf{}

* Recency guarantees on data items
	* CAP

@@

 \newcommand{\lostupdate}{$^\dagger$}
 \newcommand{\rwskew}{$^\ddagger$}
 \newcommand{\linearizable}{$^\oplus$}

\begin{table}
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (P-CI), Phantom Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), Causal
Consistency\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Read Consistency (CR)\lostupdate, Snapshot Isolation (SI)\lostupdate,
Repeatable Read (RR)\lostupdate\rwskew, Serializability
(S)\lostupdate\rwskew, Causal
Serializability\lostupdate\rwskew, Strict
Serializability\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  linearizability\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \node[draw=none,fill=none] (ici) at (1.2, 1.6) {\underline{I-CI}};
 \node[draw=none,fill=none] (pci) at (1.2, 2.4) {\underline{P-CI}};
 \node[draw=none,fill=none] (rc) at (1.2, .8) {\underline{RC}};
 \node[draw=none,fill=none] (ru) at (1.2, 0) {\underline{RU}};

 \node[draw=none,fill=none] (ta) at (0, 0) {\underline{TA}};
 \node[draw=none,fill=none] (mr) at (2.4, 0) {\underline{MR}};
 \node[draw=none,fill=none] (mw) at (3.6, 0) {\underline{MW}};
 \node[draw=none,fill=none] (wfr) at (4.9,0) {\underline{WFR}};
 \node at (6.3,0) [sticky] (ryw) {\underline{RYW}};

 \node at (3.6, 2) [sticky] (causal) {\underline{Causal}};
 \node[noha] (cs) at (2, 2.8) {CS};
 \node[noha] (rr) at (.6, 3.0) {RR};
 \node[noha] (si) at (-.4, 2.4) {SI};
 \node[noha] (cr) at (-.4, 1.2) {CR};
 \node[noha] (s) at (-.4, 3.5) {S};
 \node[noha] (causal-s) at (3.6, 3.5) {Causal S};
 \node[noha] (strict-s) at (6, 3.5) {Strict S};

 \draw [->] (ru) -- (rc);
 \draw [->] (rc) -- (ici);
 \draw [->] (ici) -- (pci);

@@

 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (mr);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (ta);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ici) -- (ta);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (pci) -- (ta);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (rc) -- (mr);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (pci) -- (mr);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ici) -- (mr);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ta) -- (ru);
 %\draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (ru) -- (causal);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (mr) -- (mw);
 %\draw[snake=coil, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (mw);
 %\draw[snake=coil, blue, segment aspect=0, segment amplitude=.75pt, segment length=2pt] (wfr) -- (ryw);

 \draw [->, red] (ici) -- (cs);
 \draw [->, red] (cs) -- (rr);
 \draw [->, red] (pci) -- (si);
 \draw [->, red] (rr) -- (s);
 \draw [->, red] (si) -- (s);
 \draw [->, red] (ta) -- (cr);
 \draw [->, red] (rc) -- (cr);
 \draw [->, red] (cr) -- (si);
 \draw [->, red] (ta) -- (rr);
 \draw [->, red] (s) -- (causal-s);
 \draw [->, red] (causal) -- (causal-s);
 \draw [->, red] (causal-s) -- (strict-s);

\end{tikzpicture}
\label{fig:hat-order}
\caption{HAT, sticky HAT (in boxes), and unavailable models (circled)
  from Table~\protect\ref{table:hatcompared} compared
  graphically. Directed edges represent ordering by model strength,
  while all underlined models can be combined (e.g., the powerset of
  the set of underlined models are all valid models). The combination
  of all underlined, non-boxed models represents maximum ACID HAT
  semantics, while the combination of underlined and boxed models
  represents maximum ACID sticky HAT models.}
\label{fig:hatcompared}
\end{figure}
